---
title: "Plots and Plots and Plots"
author: "Seth White"
date: "2021-06-08"
categories: ["R"]
tags: ["Plot"]
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Plots and Plots

For this post we are going to work through doing some of the basic plots in R. I am going to start with loading in the appropriate packages. I will be using he tidyverse set of commands which includes ggplot 

```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(cowplot)
```

## Bar Graphs

The first plot we will look at are bar graphs, we will be using the starwars data set for the bar graphs. Bar graphs are great for categorical data and getting counts of different levels of categorical data in a data set. The Starwars data set included in the dplyr package has some good examples of that

first we group and arragne the data to get a count of the different species in star wars

Then we will make a bar graph of the just the human and droids in starwars and what homeworld they are from

```{r bar graph human-droid, warning=FALSE}

starwars%>%
  select(species)%>%
  group_by(species)%>%
  count()%>%
  arrange(desc(n))


human_droid <- starwars%>%
  filter(species == "Human" | species == "Droid")%>%
  ggplot()+
  geom_bar(mapping = aes(x=homeworld, fill = species),
           position = "dodge")+
  labs(title = "World Counts in Star Wars",
       x= "Different Homeworlds",
       y = "Number of Characters from World",
       fill = "Character Species")+
  theme_minimal_hgrid()+
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.7,
                                   size = 10),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(size = 12)
        )

human_droid

```
Now we will make a second bar graph with the homeworlds of the most common species besides human and droids

```{r bar graph other species, warning=FALSE}

other_species<-starwars%>%
  group_by(species)%>%
  mutate(character_count = n())%>%
  filter(character_count<6 & character_count>1)%>%
  ggplot()+
  geom_bar(mapping = aes(x=homeworld, fill = species),
           position = "dodge")+
  labs(title = "World Counts in Star Wars",
       x= "Different Homeworlds",
       y = "Number of Characters from World",
       fill = "Character Species")+
  theme_minimal_hgrid()+
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.7,
                                   size = 10),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        plot.title = element_text(size = 12)
        )

other_species

```

Now we will combine those two bar graphs into one plot using cowplot

```{r combine bar graphs, warning=FALSE}

plot_grid(human_droid, other_species)

```

## Histograms

Now we will focus on creating some histograms of the midwest data, since that data set has a variety of numeric variables

The first histogram that we are going to make is one showing number of counties with a certain population in the midwest. There is a wide range of values in the county populations so its hard to see anything in this graph.

```{r histograms1, warning=FALSE}
midwest%>%
  ggplot()+
  geom_histogram(mapping = aes(x=poptotal, fill = state),
                 position = "dodge" )+
  labs(title = "Histogram of County Populations",
       subtitle = "In the Midwest",
       x = "Total County Population",
       y = "Number of Counties")
```

We will do the same but make a grid of histogram, one for each state in the Midwest dataset
```{r histograms1a, warning=FALSE}
midwest%>%
  ggplot()+
  geom_histogram(mapping = aes(x=poptotal, fill = state), 
                 position = "dodge")+
  labs(title = "Histogram Grid of County Populations by State",
       subtitle = "In the Midwest",
       x = "Total County Population",
       y = "Number of Counties")+
  facet_wrap(~state)
```


To try and get a little more clarity, we can separate the data into populations less than 100,000 people and look at that distribution for all 5 states
```{r histograms2, warning=FALSE}
midwest%>%
  ggplot()+
  geom_histogram(mapping = aes(x=poptotal, fill = state), 
                 position = "dodge" )+
  labs(title = "Histogram of County Populations",
       subtitle = "Less than 100,000 in Midwest",
       x = "Total County Population",
       y = "Number of Counties")+
  xlim(0,100000)
```

below we split the graph of populations below 100,000 into the different states

```{r histograms2a, warning=FALSE}
midwest%>%
  ggplot()+
  geom_histogram(mapping = aes(x=poptotal, fill = state),
                 position = "dodge")+
  labs(title = "Histogram Grid of County Populations by State",
       subtitle = "Less than 100,000 in Midwest",
       x = "Total County Population",
       y = "Number of Counties")+
  xlim(0,100000)+
  facet_wrap(~state)
```

For this last histogram we edited our graph a little bit to instead of showing counts we show the population of the top ten higest populatted counties in the midwest.

```{r histograms3, warning=FALSE}
midwest%>%
  arrange(desc(poptotal))%>%
  head(10)%>%
  ggplot()+
  geom_histogram(mapping = aes(x=county, y = poptotal, fill = state),
                 stat = "identity", 
                 position = "dodge" )+
  labs(title = "Histogram of  10 largest County Populations",
       subtitle = "In the MidWest",
       y= "Total County Population",
       x = "County Name",
       fill = "State Name")+
  theme_minimal_hgrid()+
  theme(axis.text.x = element_text(size = 7.5,
                                   angle = 45,
                                   vjust = 0.8))


```


## Box Plots

We will look at the same data with boxplots

THis first plot is a box plot for each state with the top 5 counties by population highlighted in red

```{r boxplots1, warning=FALSE}

midwest_toppop<-midwest%>%
  arrange(desc(poptotal))%>%
  head(5)

midwest%>%
  ggplot()+
  geom_boxplot(mapping = aes(x=state, 
                             y=poptotal))+
  geom_point(data=midwest_toppop, 
               mapping = aes(x=state, 
                             y = poptotal), 
               color = "red")+
  geom_label_repel(data=midwest_toppop,
                   mapping = aes(x=state, 
                                 y = poptotal, 
                                 label = county))+
  labs(title = "Boxplots of State County Populations")+
  coord_flip()
```


Next we show the same graph but exclude any counties above 500,000 people. This gives us a better look of what the population distributions actually look like

```{r boxplots2, warning=FALSE}
midwest%>%
  ggplot()+
  geom_boxplot(mapping = aes(x=state, y=poptotal))+
  labs(title = "Boxplots of State County Populations below 500,000")+
  coord_flip(ylim = c(0,500000))
```

This last plot is a new box plot based on the data points with populations above 500,000

```{r boxplots3, warning=FALSE}
midwest%>%
  ggplot()+
  geom_boxplot(mapping = aes(x=state, y=poptotal))+
  labs(title = "Boxplots of State County Populations")+
  coord_flip()+
  ylim(500000,5200000)


```



